---
title: "R preliminaries"
# subtitle: "Getting started"
author: "Mikhail Dozmorov"
institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
date: "08-25-2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px"
)
```

## R expressions, function calls, and objects

According to John Chambers, one of the creators of R’s precursor S:

- Everything that exists in R is an **object**

- Everything that happens in R is a **call to a function**

---
## Assignment operator

- We often need to save a function's result or output. For this we use the assignment operator: ` <- `, preferred over ` = `  
```{r}
scores <- mtcars
``` 
 
Now we can use `scores` as an argument to other functions. For example, compute summary statistics for each column in the data:

```{r}
summary(scores[1:4]) # First four elements
```

Use `Alt + -` (Win) or `Option + -` (Mac) in RStudio to quickly insert ` <- `

---
## Variables

- **Scalars** (0-dimensional): `a = 42`, `b = a / 7`

- **Vectors** (1-dimensional): `b = c(12, 14, 16)`
    - Access vector element as `b[2]` (returns 14)

- **Matrices** (2-dimensional):

```{r}
mtx = matrix(data = c(3, 1, 3, 2, 3, 2), ncol = 2)
mtx
```

---
## Variable names

- camelCase, snake_case, PascalCase

- Be careful not to name your variables as function names. E.g., `c` is a bad variable name because `c()` is a function for combining variables. Check its help function `?c`

- With auto-completion in RStudio, you don't need to worry about variable name length - make names that are self-explanatory

- Check [Google R Style Guide](https://google.github.io/styleguide/Rguide.html), [The Tidyverse Style Guide](http://adv-r.had.co.nz/Style.html), [Bioconductor Coding Style](https://bioconductor.org/developers/how-to/coding-style/)

.small[ https://betterprogramming.pub/string-case-styles-camel-pascal-snake-and-kebab-case-981407998841 ]

---
## Variable types

```{r}
# numeric: real or decimal numbers, sometimes referred to as “double” integer: a subset of numeric in which numbers are stored as integers
a <- 2
# character: sometimes referred to as string data, tend to be surrounded by quotes
a <- "2" 
# logical: Boolean data (TRUE and FALSE)
a <- TRUE 
```

- complex: complex numbers with real and imaginary parts (e.g., 1 + 4i)
- raw: bytes of data (machine-readable, but not human readable)

Auxillary functions

``` r
class(a)
str(a)
is.numeric() # TRUE is matches, same with is.character
as.numeric("2") # Attempt to convert types
```

---
## Factors

- Factors are how R represents categorical data
- There are two kinds of factors:
    - `factor()` - used for nominal data ("Cats", "Cats", "Dogs", "Birds")
    - `ordered()` - used for ordinal data ("First", "Second", "Second", "Third")

```{r}
factor(c("Cats", "Cats", "Dogs", "Birds"))
ordered(c("First", "Second", "Second", "Third"))
```

---
## Factors Auxillary functions

- `levels()` - get levels of a factor. Also, an argument in the `factor()` function allowing to set the order manually
- `relevel()` - reorder factor levels
- `is.factor()`, `as.factor()`

```{r}
a <- factor(c("Cats", "Cats", "Dogs", "Birds"))
a
relevel(a, ref = "Cats") 
levels(a) <- rev(levels(a))
a
```

---
## Data frames

- **Data frames**: tables or 2-dimensional arrays. Think matrices that can hold different data types
    - The column names should be non-empty
    - Columns should be the same length
    - The row names should be unique
    - The data stored in a data frame can be of numeric, factor, or character

```{r}
dat = data.frame(Column.1 = c(3, 1, 3), Column.2 = c("2", "3", "2"))
dat
```

---
## Data frames Auxillary functions

```{r}
dim(dat)
nrow(dat)
ncol(dat)
length(dat)
colnames(dat)
rownames(dat)
```

---
## Subsetting elements in a data frame

```{r}
dat[3, 2]          # [] contain row/column indices. 
dat[3, "Column.2"] # Address by column name 
dat$Column.2[3]    # Use $ shortcut to access column by name
```

```{r}
# Compare column classes
class(dat$Column.1)
class(dat$Column.2)
```

---
## Subsetting elements in a data frame

```{r}
index <- dat$Column.1 > 1
index
dat[index, ]
```

```{r}
index <- dat$Column.1 == 1
dat[index, "Column.2"]
dat[index, "Column.2", drop = FALSE]
```


---
## Inspecting data.frame objects

There are several built-in functions that are useful for working with data frames.

* Content:
    * `head()`: shows the first few rows
    * `tail()`: shows the last few rows

* Size:
    * `dim()`: returns a 2-element vector with the number of rows in the first element, and the number of columns as the second element (the dimensions of the object)
    * `nrow()`: returns the number of rows
    * `ncol()`: returns the number of columns

---
## Inspecting data.frame objects

* `colnames()` (or just `names()`): returns the column names

* `str()`: structure of the object and information about the class, length and content of each column

* `summary()`: works differently depending on what kind of object you pass to it. Passing a data frame to the `summary()` function prints out useful summary statistics about numeric column (min, max, median, mean, etc.)
