---
title: "R preliminaries"
# subtitle: "Getting started"
author: "Mikhail Dozmorov"
institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
date: "08-30-2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px"
)
```
```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("editable", "panelset"))
```

## Useful ways of getting data in and out of R

- Base functions: `read.table`, `read.csv`, `write.table`, `write.csv`

- Tidyverse way, `readr` package: `read_table`, `read_csv`, `read_tsv`, `write_csv` ...

- For fixed-width files, use `read.fwf` or `readr::read_fwf` funcitons

- For reading/writing Excel files, use `readxl` and `writexl` packages, `read_xlsx`, `write_xlsx` functions
    - Remember that `.csv` is the preferred text-based format that opens in Excel

.small[https://readr.tidyverse.org/

https://readxl.tidyverse.org/

https://CRAN.R-project.org/package=writexl]

---
## The stringsAsFactors curse

- When creating data frames with `data.frame()` or reading data with `read.table()`, strings automatically converted to factors

- This behind-the-scenes factor conversion can lead to unpredictable behaviors

- Use `as.is = TRUE` in `read.table()` to avoid such conversion

- Better yet, set `options(stringsAsFactors = FALSE)` at the beginning of your script files

.small[https://developer.r-project.org/Blog/public/2020/02/16/stringsasfactors/]

---
## Save/load R objects

- `save()`, `load()` - saves/loads R objects to the specified file
``` r
x <- stats::runif(20)
y <- list(a = 1, b = TRUE, c = "oops")
save(x, y, file = "xy.rda")
load(file = "xy.rda")
```

- `saveRDS()`, `readRDS()` - saves/loads a _representation_ of the object
``` r
x <- stats::runif(20)
saveRDS(x, file = "x.rds")
x2 <- readRDS(file = "x.rds")
identical(x, x2, ignore.environment = TRUE)
```

.small[https://fromthebottomoftheheap.net/2012/04/01/saving-and-loading-r-objects/]

---
## R datasets

R contains many datasets (stored as data frames) that are built-in to the software

```{r}
data() # All built-in datasets
# ?trees
data(trees) # Load a particular one
head(trees)
```

---
## Accessing data in datasets

```{r}
attach(trees)   # You can make R find variables in any data frame by adding the data frame to the search path
search()        # .GlobalEnv is your workspace and the package quantities are libraries
detach(trees)   # To remove an object from the search path, use the detach()
with(trees, mean(Height)) # Evaluate an R expression in an environment constructed from data, possibly modifying (a copy of) the original data
```

`attach()` can cause name overloads and other serious issues. Avoid it

---
## Summary statistics

- Simple statistical functions: `count()`, `min()`, `max()`, `mean()`, `median()`, `sd()`, `cor()`, `summary()`)

- These, and many other functions, have settings to properly handle NAs, e.g., `mean(x, trim = 0, na.rm = FALSE, ...)`

- `complete.cases()` on a matrix/data frame returns row-wise logical with TRUE for rows without NAs

- `unique()` - unique elements in a vector. Combine with `length()` to get the number of unique elements

- `table()` - contingency table for a vector (the number of elements per unique level)

---
## Summary statistics

```{r}
data(mtcars)    # simple summary 
# ?mtcars
head(mtcars)
```

```{r}
mean(mtcars$mpg) # Try median, sd, var, min, max
summary(mtcars$mpg)
```

---
## Summary statistics

```{r}
quantile(mtcars$mpg, probs = c(.20, .80))
cor(mtcars$mpg, mtcars$hp) # sample correlation coeficient
```

```{r}
table(mtcars$cyl)
table(mtcars$cyl)/length(mtcars$cyl) # normalized by the total number of observations = 32
```

---
## Control structures inside R/functions

- `if, else`
- `for`
- `while`
- `repeat`
- `break`
- `next`

---
## If-else statement

Conditional code execution

``` r
if (condition) {
  # do something
} else {
  # do something else
}
```

- `==`: Equal to
- `!=`: Not equal to
- `>`, `>=`: Greater than, greater than or equal to
- `<`, `<=`: Less than, less than or equal to

```{r}
x <- 1:15
if (sample(x, 1) <= 10) {
  print("x is less than 10")
} else {
  print("x is greater than 10")
}
```

---
## For loop

Repetitive code execution

```{r}
for (i in 1:5) {
  cat(i)
}
```

Compare with

```{r}
for (i in 1:5) {
  print(i)
}
```

---
## More uses of For loops

```{r}
x <- c("apples", "oranges", "bananas", "strawberries")

for (i in x) {
  cat(i); cat(" ")
}
```

```{r}
for (i in 1:4) {
  cat(x[i]); cat(" ")
}
```

```{r}
for (i in seq(x)) {
  cat(x[i]); cat(" ")
}
```

---
## Nested For loops

```{r}
m <- matrix(1:10, 2)
m
for (i in seq(nrow(m))) {
  for (j in seq(ncol(m))) {
    print(m[i, j])
  }
}
```

---
## while, repeat loops

```{r}
i <- 1
while (i < 10) {
  print(i)
  i <- i + 1
} # Be sure there is a way to exit out of a while loop
```

``` r
repeat {
  # simulations; generate some value have an expectation if within some range,
  # then exit the loop
  if ((value - expectation) <= threshold) {
    break
  }
}
```

---
## Combine any statements/functions

```{r}
for (i in 1:20) {           
  if (i%%2 == 1) {
    next                # skip printing over odd numbers
  } else {
    print(i)
  }
}
```

---
## Vectorized operation

Many operations in R are already vectorized, making code more efficient, concise, and easier to read

```{r}
x <- 1:4; y <- 6:9
x
y
x * y
x / y
```

---
## Manipulating vectors

```{r}
ages <- c(40, 50, 60, 70, 80)
# add a value to end of vector
ages <- c(ages, 90) 
# add value at the beginning
ages <- c(30, ages)
# extracting second value
ages[2]
# excluding second value
ages[-2]
# extracting first and third values
ages[c(1, 3)] 
```

---
## `apply` family of functions 

Writing for, while loops in R are inefficient, and we want to vectorize computation in R.

- `apply()` - apply a function over the margins of an array

- `lapply()` - loop over a list and evaluate a function on each element

- `sapply()` - same as lapply but try to simplify results, if the result is a list where every element is length 1, then a vector is returned

- `mapply()` - multivariate version of lapply

- `tapply()` -  apply a function over subsets of a vector

---
## apply examples

```{r}
x <- 1:4
lapply(x, runif)
```

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1))
sapply(x, mean)
```

---
## apply examples

```{r}
#If the result is a list where every element is a vector of the same length (> 1), a matrix is returned.
x <- list(rnorm(100), runif(100), rpois(100, 1))
sapply(x, quantile, probs = c(0.25, 0.75))
```

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, sum)
apply(x, 2, mean)
```

---
## apply examples

For sums and means of matrix dimensions, we have some shortcuts

```{r}
rowSums  = apply(x, 1, sum)
rowMeans = apply(x, 1, mean)
colSums  = apply(x, 2, sum)
colMeans = apply(x, 2, mean)
```

Check `?rowSums` help on these base R functions

---
## tapply

Apply a function to each cell of a ragged array, that is, to each (non-empty) group of values given by a unique combination of the levels of certain factors.

``` r
function (X, INDEX, FUN = NULL, ..., default = NA, simplify = TRUE)
X is a vector
INDEX is a factor or a list of factors (or else they are coerced to factors)
FUN is a function to be applied
... contains other arguments to be passed FUN
simplify, should we simplify the result?
```

```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
tapply(x, f, mean)
```

---
## mapply

.panelset[
.panel[.panel-name[mapply]
mapply is a multivariate version of sapply. mapply applies FUN to the first elements of each ... argument, the second elements, the third elements, and so on. Arguments are recycled if necessary.

``` r
function (FUN, ..., MoreArgs = NULL, SIMPLIFY = TRUE, USE.NAMES = TRUE)
FUN is a function to apply
... contains arguments to apply over
MoreArgs is a list of other arguments to FUN.
SIMPLIFY indicates whether the result should be simplified
```
]
.panel[.panel-name[Example 1]
```{r}
mapply(rep, 1:4, 4:1)
```
]
.panel[.panel-name[Example 1]
```{r}
mapply(rnorm, mean = 1:3, sd = 1:3, n = seq(5, 15, by = 5))
```
]
]

---
## Formatting

```{r}
round(c(3.14159, 2.71828), digits = 2)
```

```{r}
formatC(c(3.14159, 2.71828), digits = 2, format = "e")
```

---
## Handling NAs

- NAs create problems in calculations. Many functions have built-in mechanism to handle NAs. Check `?mean` and the 'na.rm' argument

- Check for presence of NAs in your data using `is.na()`

```{r}
vec_with_NAs <- c(1, 2, NA, 3)
sum(is.na(vec_with_NAs))
```

- Remove all **rows** containing NAs from a data frame

```{r eval=FALSE}
dat <- dat[complete.cases(dat), ]
```

---
## Modifying vector elements

.can-edit[
```{r}
a <- c("Heineken", "Tuborg", "Carlsberg")
paste("Beer", a)
paste("Beer", a, sep = ":")
paste("Beer", a, collapse = ":")
```

```{r}
paste0("e", "yes")
```
]

---
## Modifying vector elements

```{r}
a <- c("but", "cut")
sub("u", "a", a)
```

```{r}
a <- c("Column.1", "Column.2")
strsplit(a, ".", fixed = TRUE)
sapply(a, function(x) strsplit(x, ".", fixed = TRUE)[[1]][2])
```

