---
title: "Useful stats functions and plots, text manipulation, regular expressions"
# subtitle: "Getting started"
author: "Mikhail Dozmorov"
institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
date: "09-08-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px"
)
```

## Empirical Cumulative Distribution Function

```{r fig.height=5}
set.seed(4)
Fn <- ecdf(x <- rnorm(100)) # ?ecdf()
curve(Fn) # Try plot(Fn)
```

---
## Quantile-Quantile Plots

- `qqnorm()`, `qqline()`, `qqplot()` - distribution comparison plots

```{r fig.height=5}
data(trees) # load data to global environment
attach(trees)
qqnorm(Height) # A normal QQ plot 
```

---
## Scatterplot Matrices

- `pairs()` - pair-wise plot of multivariate data

```{r fig.height=5}
data(trees)
pairs(trees) 
```

---
## Random samples and permutations

```{r}
sample(1:10)
sample(letters, 5)
sample(1:10, replace = TRUE)
```

```{r}
set.seed(1)
```

---
## Examples of density functions

Prefix each R distribution name with + ‘d’ for the density or mass function, + ‘p’ for the CDF, + ‘q’ for the percentile function (also called the quantile), + ‘r’ for the generation of pseudorandom variables

| Function | Distribution |
|:--------:|:------------:|
| dnorm    | Normal       |
| dpois    | Poisson      |
| dbinom   | Binomial     |
| dchisq   | Chi-squared  |
| dt       | Student’s t  |
| dunif    | Uniform      |


---
```{r}
x=rnorm(100)
y=rnorm(100)
plot(x, y)
```


```{r}
qnorm(.75,mean=10,sd=2) # 3rd quartile of N(mu = 10,sigma = 2)
qnorm(c(0.05, 0.10, 0.20, 0.95),mean=10,sd=2)
qt(.95,df=20) # 95th percentile of t(20)
```

---
```{r}
x<-rchisq(100,1)
plot(x)
hist(x)
```

```{r}
x<-dbinom(3:10,size=10,prob=.25)   # P(X=3) for X ~ Bin(n=10, p=.25)

barplot(x)
plot(x) 
plot(0:10, dbinom(0:10, size=10, prob=.25), type = "h", lwd = 30)
plot(3:10, x, type = "h", lwd = 30, main = "Binomial Probabilities w/ n = 10, p = .25", ylab = "p(x)") # which is gives the histogram-like vertical lines 
# lwd option (the default width is 1) controls line thickness
```

---
```{r fig.height=5}
dpois(0:2, lambda=4) # P(X=0), P(X=1), P(X=2) for X ~ Poisson
x<- dpois(0:20, lambda=4)
barplot(x)
# plot(x)
```

---
```{r fig.height=5}
pbinom(3,size=10,prob=.25) # P(X <=3) in the above distribution
x<- pbinom(3:10,size=10,prob=.25)
plot(x)
```

---
## More useful stats functions

```{r}
lm(Sepal.Length~Sepal.Width, data=iris) # simple linear regression
```

```{r}
glm(ifelse(Species=="setosa",1,0)~Sepal.Width, family="binomial",data=iris)  # logistic regression
```

---
## More useful stats functions

```{r}
t.test(iris$Sepal.Length,iris$Petal.Length)
```

```{r}
aov(Sepal.Length~Species,data=iris)
```

---
## More useful stats functions

```{r}
chisq.test(iris$Petal.Length,iris$Species)
```

```{r}
fisher.test(mtcars$gear, mtcars$carb)
```

---
## Regression models

Regression models can be used to estimate how the expected value of a dependent variable changes as independent variables change.

In R, regression formulas take this structure:

```
## Generic code
[response variable] ~ [indep. var. 1] +  [indep. var. 2] + ...
```

Notice that a tilde, ~, is used to separate the independent and dependent variables and that a plus sign, +, is used to join independent variables. This format mimics the statistical notation:

$Y_i \sim X_1 + X_2 + X_3$

---
## Conventions for linear models

| Convention	| Meaning |
|:-----------:|:-------:|
| I()	| evaluate the formula inside I() before fitting (e.g., I(x1 + x2)) |
| :	| fit the interaction between x1 and x2 variables |
| *	| fit the main effects and interaction for both variables (e.g., x1*x2 equals x1 + x2 + x1:x2)|
| .	| include as independent variables all variables other than the response (e.g., y ~ .) |
| 1	| intercept (e.g., y ~ 1 for an intercept-only model) | 
| -	|do not include a variable in the data frame as an independent variables (e.g., y ~ . - x1); usually used in conjunction with . or 1 | 

---
## Linear models

To fit a linear model, you can use the function `lm()`. This function is part of the `stats` package, which comes installed with base R

```{r}
mod <- lm(mpg ~ hp, data = mtcars)
# Check class() and str() of the mod object
```

This previous call fits the model:

$Y_{i} = \beta_{0} + \beta_{1}X_{1,i} + \epsilon_{i}$

.small[ [INTRODUCTION TO LINEAR MIXED MODELS](https://ourcodingclub.github.io/tutorials/mixed-models/) ]

---
## Manipulating the `lm` object

| Function	| Description |
|:---------:|:-----------:|
| summary	  | Get a variety of information on the model, including coefficients and p-values for the coefficients |
| coefficients	| Pull out just the coefficients for a model |
| fitted	| Get the fitted values from the model (for the data used to fit the model) | 
| plot	| Create plots to help assess model assumptions
| residuals	| Get the model residuals |

```{r}
class(mod)
```


---
## plot(mod)

```{r echo=FALSE, fig.height=4}
plot(mod)
```

---
## Manipulating the `lm` object

```{r fig.height=4}
mod_coef <- coefficients(mod)
library(ggplot2)
ggplot(mtcars, aes(x = hp, y = mpg)) + 
  geom_point(size = 1) + 
  xlab("Miles/(US) gallon") + ylab("Gross horsepower") + 
  geom_abline(aes(intercept = mod_coef[1],
                  slope = mod_coef[2]), col = "red")
```

---
## Working with text

- The `grep()` function takes as parameters the pattern and a character vector as the data to search through for the pattern. Parameters:
    - `ignore.case = FALSE` - by default it is case sensitive
    - `value = FALSE` - by default returns vector with index values of match; otherwise returns the values
    - `fixed = FALSE` - by default treats pattern as regular expression; otherwise will match exact
    - `invert = FALSE` - by default matches the pattern; otherwise returns what is not matched

```{r}
strings <- c('abcd', 'dabc', 'abcabc')
pattern <- '^abc'
print (grep(pattern, strings))
```

.small[`grepl()` - grep logical, returns a vector of the same length as a string, with TRUE/FALSE pattern matching]

---
## Regular expressions

Some useful regular expression operators include:

| Expression | Description                                                                                                                                     |
|:----------:|-------------------------------------------------------------------------------------------------------------------------------------------------|
|     []     | Matches a set. [abc] matches a, b, or c. [a-zA-Z] matches any letter. [0-9] matches any number. “^” negates a set, [^abc] matches d, e, f, etc. |
|      ^     | Starting position anchor. ^abc finds lines starting with abc                                                                                    |
|      "$"     | Ending position anchor. xyz"$" finds lines ending with xyz. Remove quotes                                                                                        |
|      \\     | Escape symbol, to find special characters. \\\* will find \*. \\n matches new line character, \\t – tab character                                  |
|      \*     | Match the preceding element zero or more times. a\*b matches ab, aab, aaab, etc.                                                                 |

---
## Extended regular expressions

| Expression | Description                                                                    |
|:----------:|--------------------------------------------------------------------------------|
|      ?     | Matches the preceding element zero or one time. a*b matches b, ab, but not aab |
|      +     | Matches the preceding element one or more times. a+b matches ab, aab, etc.     |
|      &#124;     | OR operator. "abc&#124;def" matches abc or def                                      |
|   .	       | Any character |

---
## Special characters

| Expression | Description  |
|:----------:|--------------|
|   \\n       | Newline      |
|   \\r       | Return       |
|   \\t       | Tab          |

https://www.regular-expressions.info/refcharacters.html

Online preview: https://regexr.com/

Play Regex Golf: https://alf.nu/RegexGolf
