---
title: "Code optimization best practices"
# subtitle: "Getting started"
author: "Mikhail Dozmorov"
institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
date: "09-27-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px",
  code_font_size = "26px"
)
```

## Timing

Use `system.time()` functions to measure the time of execution.

``` r
> # make a function
> myFun <- function(x) {
+    y = vector(length=x)
+    for (i in 1:x) y[i]=i/(i+1)
+    y
+ }


> # execute the function, measuring the time of the execution
> system.time( myFun(100000) )
   user  system elapsed
  0.107   0.002   0.109
```

---
## Memory

`utils::object.size()` - Report the Space Allocated for an Object

```{r eval=FALSE}
sl <- object.size(seq(1:10^6))

print(sl) ## 4000048 bytes
print(sl, units = "auto") ## 3.8 Mb
print(sl, units = "auto", standard = "IEC") ## 3.8 Mib
print(sl, units = "auto", standard = "SI") ## 4 MB
```

Alternatively, `pryr::object_size()` function to measure memory footprint of R objects

``` r
> library(pryr)
> object_size(USArrests)
5.23 kB
```

---
## Code speedup: Use vectors

``` r
> # using loops
> g1 <- function(x) {
+    y = vector(length=x)
+    for (i in 1:x) y[i]=i/(i+1)
+    y
+ }

> # execute the function 
> system.time( g1(100000) )
   user  system elapsed
  0.107   0.002   0.109
```

---
## vs.

``` r
> # using vectors
> x <- (1:100000)
> g2 <- function(x) {
+    x/(x+1)
+ }

> # execute the function
> system.time( g2(x) )
 user  system elapsed
  0.002   0.000   0.003
```

---
## Pre-allocate arrays

``` r
> vec1<-NULL  


> # execute the command
> system.time(
+ for(i in 1:100000) 
+ vec1 <- c(vec1,mean(1:100)))
    user  system elapsed
  58.181   0.193  58.417
```

---
## vs.

``` r
> vec2 <- vector(
+ mode=“numeric”,length=100000)

> # execute the command
> system.time(
+ for(i in 1:100000) 
+ vec2[i] <- mean(1:100))
   user  system elapsed
  2.324   0.063   2.388
```

---
## Use optimized R-functions

- `rowSums()`,  `rowMeans()`,  `table()`,  etc.

``` r
> matx <- matrix(rnorm(1000000),100000,10)

> # execute the command
> system.time(apply(matx,1,mean))
    user  system elapsed
   2.686   0.057   2.748
```

---
## vs.

``` r
> matx <- matrix(rnorm(1000000),100000,10)

> # execute the command
> system.time(rowMeans(matx))
   user  system elapsed
  0.013   0.000   0.014
```

---
## Parallelization: parallel

- The `parallel` R core package has `mclapply()` function for multi-core lapply (Mac, Linux)
- Implements functionality previously provided by the `multicore` and `snow` packages

```{r}
library(parallel)
library(MASS)

starts <- rep(100, 40)
fx <- function(nstart) kmeans(Boston, 4, nstart=nstart)
system.time( results <- lapply(starts, fx) )
```

---
## Parallelization: parallel

- The `parallel` R core package has `mclapply()` function for multi-core lapply (Mac, Linux)

```{r}
(numCores <- detectCores())
system.time( results <- mclapply(starts, fx, mc.cores = numCores) )
```

.small[ https://nceas.github.io/oss-lessons/parallel-computing-in-r/parallel-computing-in-r.html ]

---
## Parallelization methods: foreach/doParallel

- Combined with `doParallel` functionality, the `foreach` R package enables parallelized for loops

```{r message=FALSE}
library(foreach)
library(doParallel)
registerDoParallel(numCores)  # use multicore
foreach (i=1:3) %dopar% {
  sqrt(i)
}
```

Used by over 700 packages on CRAN and Bioconductor

.small[ https://CRAN.R-project.org/package=foreach ]

---
## Parallelization methods: BiocParallel

- Unified interface to the methods for parallel evaluation

- The `bplapply()` function performs parallel lapply

```{r}
library(BiocParallel)
system.time( results <- bplapply(starts, fx) )
```

.small[ http://lcolladotor.github.io/2016/03/07/biocparallel/ ]

---
## Rcpp = R and C++

- R is a high-level _interpreted_ language

- C/C++ are low-level _compiled_ languages

- C is approximately more than 50X times faster than R

- R is much better for prototyping - one line of code in R is typically many lines of code in C/C++

- `Rcpp` was created by Dirk Eddelbuettel and Romain Francois in 2011. Permits direct interchange of rich R objects between R and C++

.small[ http://adv-r.had.co.nz/Rcpp.html

http://dirk.eddelbuettel.com/code/rcpp.html ]

---
## Code profiling

Profiling is a tool, which can be used to find out how much time is spent in each function. Code profiling can give a way to locate those parts of a program which will benefit most from optimization.

- `Rprof()` – turn profiling on
- `Rprof(NULL)` – turn profiling off
- `summaryRprof("Rprof.out")` – Summarize the output of the `Rprof()` function to show the amount of time used by different R functions.

``` r
> summaryRprof("bmslow.out")
$by.self
         self.time self.pct total.time total.pct
"cbind"     400.52    99.39     400.52     99.39
"rnorm"       1.70     0.42       1.70      0.42
"bmslow"      0.74     0.18     402.96    100.00
```


---
## Code profiling

- `microbenchmark` - Accurate Timing Functions. Provides infrastructure to accurately measure and compare the execution time of R expression

- `profvis` -  Interactive Visualizations for Profiling R Code Overview, 

- `bench` - High Precision Timing of R Expressions

.small[ https://CRAN.R-project.org/package=microbenchmark

https://rstudio.github.io/profvis/

http://r-lib.github.io/bench ]

---
## R goodies

- `skimr` - A frictionless, pipeable approach to dealing with summary statistics, [https://github.com/ropenscilabs/skimr](https://github.com/ropenscilabs/skimr)

- `data.table` - fast data reading, subsetting, aggregating, summarizing,  [https://github.com/Rdatatable/data.table/wiki/Getting-started](https://github.com/Rdatatable/data.table/wiki/Getting-started)

- [dtplyr](https://dtplyr.tidyverse.org/) - The data.table backend to dplyr. [Overview](https://www.business-science.io/code-tools/2021/04/13/dtplyr-datatable-dplyr-backend.html), [Tweet](https://twitter.com/mdancho84/status/1441776263702712320?s=20)

- Whenever you get a strange execution error it is sometimes helpful to show the history of all the function calls leading to that error. This is done by typing `traceback()` at the command prompt


