---
title: "Classes and Methods in R"
subtitle: "https://github.com/rdpeng/CourseraLectures/blob/master/classes-methods.Rnw"
author: "Modified after Roger Peng's lecture"
institute: "https://youtu.be/93N0HdoZW9g"
# date: "`r Sys.Date()`"
date: "10-11-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px",
  code_font_size = "26px"
)
```

## Classes and Methods

- A system for doing object oriented programming

- R is both interactive *and* has a system for object orientation.
    - Other languages which support OOP (C++, Java, Lisp, Python, Perl)
generally speaking are not interactive languages

<!--
-  In R much of the code for supporting classes/methods is written
  by John Chambers himself (the creator of the original S language)
  and documented in the book _Programming with Data: A Guide to the S
  Language_
-  A natural extension of Chambers' idea of allowing someone to
  cross the user $\longrightarrow$ programmer spectrum
-->

- Object oriented programming is a bit different in R than it is
  in most languages - even if you are familiar with the idea, you
  may want to pay attention to the details

---
## Two styles of classes and methods

S3 classes/methods

- Essentially a `list` with a `class` attribute on it
- Informal, you can assign any class to any list, which is nonsense
- Sometimes called *old-style* classes/methods

S4 classes/methods

- More formal and rigorous
- Included with S-PLUS 6 and R  1.4.0 (December 2001)
- Also called *new-style* classes/methods


---
## Two worlds living side by side

-  For now (and the forseeable future), S3 classes/methods and S4
  classes/methods are separate systems (but they can be mixed to some
  degree).
  
- Each system can be used fairly independently of the other.

- Developers of new projects (you!) are encouraged to use the S4
style classes/methods.

<!--
- But many developers still use S3 classes/methods because they
are "quick and dirty" (and easier).
- In this lecture we will focus primarily on S4 classes/methods
-  The code for implementing S4 classes/methods in R is in the
  **methods** package, which is usually loaded by default (but
  you can load it with `library(methods)` if for some reason it
  is not loaded)
-->

---
## Object Oriented Programming in R

-  A *class* is a description of an thing. A class can be
  defined using `setClass()` in the **methods** package.
  
- An *object* is an instance of a class. Objects can be created
using `new()`.

-  A *method* is a function that only operates on a certain
  class of objects.

-  A generic function is an R function which dispatches methods.  A
  generic function typically encapsulates a "generic" concept
  (e.g. `plot`, `mean`, `predict`, ...)

<!-- The generic function does not actually do any computation. -->

- A *method* is the implementation of a generic function for an
object of a particular class.




---
## Things to look up

- The help files for the `methods` package are extensive - do read
them as they are the primary documentation

- You may want to start with `?Classes` and `?Methods`

- Check out `?setClass`, `?setMethod`, and `?setGeneric`

- Use `class?` prefix for help on specific classes, e.g., `class?GRanges`

<!-- Some of it gets technical, but try your best for now - it will make
sense in the future as you keep using it. 

- Most of the documentation in the **methods** package is
  oriented towards developers/programmers as these are the primary
  people using classes/methods -->

---
## Classes

All objects in R have a class which can be determined by the class
function

```{r examplesclasses}
class(1)
class(TRUE)
class(rnorm(100))
class(NA)
class("foo")
``` 



---
## Classes (cont'd)

Data classes go beyond the atomic classes
```{r lmclass}
x <- rnorm(100)
y <- x + rnorm(100)
fit <- lm(y ~ x)  ## linear regression model
class(fit)
names(fit)
isS4(fit)
``` 

---
## Classes (cont'd)

```{r}
getClass("lm")
```

---
## Generics/Methods in R

- S4 and S3 style generic functions look different but conceptually,
they are the same (they play the same role).

- When you program you can write new methods for an existing generic OR
create your own generics and associated methods.

-  Of course, if a data type does not exist in R that matches your
  needs, you can always define a new class along with generics/methods
  that go with it


---
##An S3 generic function (in the `base` package)

The `mean` function is generic
```{r printmean}
mean
``` 

So is the `print` function
```{r printprint}
print
``` 


---
##S3 methods
```{r methodsmean}
methods("mean")
``` 


---
##An S4 generic function (from the `methods` package)
The S4 equivalent of `print` is `show`  
```{r showmethod}
show
``` 

The `show` function is usually not called directly (much like
`print`) because objects are auto-printed

---
##S4 methods


Think of S4 methods as simple functions that perform certain task depending on class

```
mimicMethod <- function(x) {
    if (is(x, "matrix"))
        method1(x)
    if (is(x, "data.frame"))
        method2(x)
    if (is(x, "IRanges"))
        method3(x)
}
```


<!--
There are many different methods for the `show` generic
function
showMethods("show")
-->

---
## Generic/method mechanism

The first argument of a generic function is an object of a particular
class (there may be other arguments)
- The generic function checks the class of the object.

- A search is done to see if there is an appropriate method for
that class.

- If there exists a method for that class, then that method is
called on the object and we're done.

- If a method for that class does not exist, a search is done to see
if there is a default method for the generic. If a default exists,
then the default method is called.

- If a default method doesn't exist, then an error is thrown.


---
## Examining Code for Methods
Examining the code for an S3 or S4 method requires a call to a special
function

-  You cannot just print the code for a method like other
functions because the code for the method is usually hidden.

-  If you want to see the code for an S3 method, you can use the function
`getS3method`.
    -  The call is `getS3method(<generic>, <class>)`

-  For S4 methods you can use the function `getMethod`
    -  The call is `getMethod(<generic>, <signature>)` (more
  details later)

---
## Examining Code for Methods

```
> showMethods("mean")
Function: mean (package base)
x="ANY"
x="AtomicList"
x="CompressedIntegerList"
x="CompressedLogicalList"
x="CompressedNumericList"
x="CompressedRleList"
x="Rle"
x="Views"
```

---
##S3 Class/Method: Example 1
 What's happening here?
```{r meanexample}
set.seed(2)
x <- rnorm(100)
mean(x)
``` 

- The class of x is `numeric`
- But there is no mean method for `numeric` objects!
- So we call the default function for `mean`.


---
##S3 Class/Method: Example 1
```{r showmean}
getS3method("mean", "default")
``` 

---
##S3 Class/Method: Example 2
  What happens here?
```{r dataframemean}
set.seed(3)
df <- data.frame(x = rnorm(100), y = 1:100)
sapply(df, mean)
```   

- The class of df is `data.frame`; in a data frame each column can be
an object of a different class
- We `sapply` over the columns and call the `mean` function
- 
In each column, `mean` checks the class of the object and
dispatches the appropriate method.
-  Here we have a `numeric` column and an `integer`
  column; in both cases `mean` calls the default method



---
## Calling Methods

- Some methods are visible to the user (i.e. `mean.default`)

- You should **never** call methods directly

- Use the generic function and let the method be dispatched automatically


---
##S3 Class/Method: Example 3
  The `plot` function is generic and its behavior depends on the
  object.
```{r plotdefault, fig.height=4, fig.width=4}
set.seed(10)
x <- rnorm(100)
plot(x)
``` 


---
##S3 Class/Method: Example 3
  For time series objects, `plot` connects the dots
```{r plotts, fig.height=3, fig.width=12}
set.seed(10)
x <- rnorm(100)
y <- x + rnorm(100)
fit <- lm(y ~ x)  ## linear regression model
par(mfrow = c(1, 4))
plot(fit)
``` 


---
## Write your own methods!
If you write new methods for new classes, you'll probably end up
writing methods for the following generics:

- print/show
- summary
- plot

There are two ways that you can extend the R system via classes/methods

-  Write a method for a new class but for an existing generic
  function (i.e. like `print`)
-  Write new generic functions and new methods for those generics




---
## S4 Classes
Why would you want to create a new class?

- To represent new types of data (e.g. gene expression, space-time,
hierarchical, sparse matrices)

- New concepts/ideas that haven't been thought of yet (e.g. a fitted
point process model, mixed-effects model, a sparse matrix)

- To abstract/hide implementation details from the user

<!-- I say things are `new` meaning that R does not know about them (not
that they are new to the statistical community).-->


---
## S4 Class/Method: Creating a New Class
  A new class can be defined using the `setClass` function

-  At a minimum you need to specify the name of the class

-  You can also specify data elements that are called *slots*

-  You can then define methods for the class with the
  `setMethod` function
  
-  Information about a class definition can be obtained with the
  `showClass` function


---
##S4 Class/Method: Polygon Class
Creating new classes/methods is usually not something done at the
console; you likely want to save the code in a separate file
```{r createpolygonclass}
setClass("polygon",
         representation(x = "numeric",
                        y = "numeric"))
``` 
The slots for this class are `x`    and `y`. The slots for an
S4 object can be accessed with the `@` operator.

You should never have to access slots directly. You should get/set data out of the class slot using the "accessor" functions (getters and setters).

---
##S4 Class/Method: Polygon Class

- A `plot` method can be created with the `setMethod`
function.
    - Specify a generic function(`plot`), and a class *signature*.

<!--  A signature is a character vector indicating the classes of
  objects that are accepted by the method. In this case, the
  `plot` method will take one type of object--a `polygon`
  object. -->

```{r createplotmethod}
setMethod("plot", "polygon",
          function(x, y, ...) {
                  plot(x@x, x@y, type = "n", ...)
                  xp <- c(x@x, x@x[1])
                  yp <- c(x@y, x@y[1])
                  lines(xp, yp)
          })
``` 

Note that the slots of the polygon (the x- and y-coordinates) are
accessed with the `@` operator.

If things go well, you will not get any messages or errors and nothing
useful will be returned by either `setClass` or `setMethod`.

---
##S4 Class/Method: Polygon Class
After calling `setMethod` the new `plot` method will be
added to the list of methods for `plot`.
```{r showMethods}
showMethods("plot")
```   

Note that the signature for class `polygon` is listed.  The
method for `ANY` is the default method and it is what is called
when now other signature matches

---
##S4 Class/Method: Polygon class
```{r showplotpolygon, fig.height=5, fig.width=5}
p <- new("polygon", x = c(1, 2, 3, 4), y = c(1, 2, 3, 1))
plot(p)
``` 

---
## Class inheritance

- Class inheritance is used when you define a new class which “is almost like this other class but with a little twist”. 

- For example `ExpressionSet` inherits from `eSet`, and when you look at the class definition you cannot easily see a difference. The difference is that `ExpressionSet` is meant to contain expression data and has the `exprs()` accessor.


---
## References

- https://github.com/MalteThodberg/S4-Bioconductor - a collection of references by Malte Thodberg

- https://github.com/nullranges/nullranges - an R package for generation of null ranges via bootstrapping or covariate matching

- [Reference classes](http://adv-r.had.co.nz/R5.html)